---
title: "SCAVENGE-vignette"
author: "Fulong Yu"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SCAVENGE-vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
## Overview

This vignette covers the main function and workflow of SCAVENGE.
The standard processed input data including fine-mapped variants and single-cell epigenomic profiles. For fine-mapped variants of the trait of interest, we typically need information of genomic locations of variants and their corresponding posterior propability of causality. A peak-by-cell matrix of scATAC-seq profiles is needed. To walk through the workflow of SCAVENGE, we provided a blood cell trait of monocyte count and a 10X PBMC dataset as an example. 


### Load required packages

```{r, message = FALSE, warning = FALSE}
library(SCAVENGE)
library(chromVAR)
library(gchromVAR)
library(BuenColors)
library(SummarizedExperiment)
library(data.table)
library(BiocParallel)
library(BSgenome.Hsapiens.UCSC.hg19)
library(dplyr)
library(igraph)

set.seed(9527)
```

### Load example data
The PBMC data was processed using [ArchR](https://www.archrproject.com/) package. The peak-by-cell count matrix and corresponding meta data were extracted and stored in a [RangedSummarizedExperiment](https://bioconductor.org/packages/3.7/bioc/vignettes/SummarizedExperiment/inst/doc/SummarizedExperiment.html) object (for more details please follow our paper).
```{r message=TRUE, warning=FALSE, cache=FALSE}
trait_file <- paste0(system.file('extdata', package='SCAVENGE'), "/mono.PP001.bed")
pbmc5krda <- paste0(system.file('rda', package='SCAVENGE'), "/pbmc5k_SE.rda")
load(pbmc5krda)
```

### gchromVAR analysis
```{r, message = TRUE, warning = FALSE, cache = TRUE}
SE_pbmc5k <- addGCBias(SE_pbmc5k, genome = BSgenome.Hsapiens.UCSC.hg19)
SE_pbmc5k_bg <- getBackgroundPeaks(SE_pbmc5k, niterations=200)
trait_import <- importBedScore(rowRanges(SE_pbmc5k), trait_file, colidx=5)
SE_pbmc5k_DEV <- computeWeightedDeviations(SE_pbmc5k, trait_import, background_peaks = SE_pbmc5k_bg)
```

**Reformat results**
```{r, message = TRUE, warning = FALSE, cache = TRUE}
z_score_mat <- data.frame(colData(SE_pbmc5k), z_score=t(assays(SE_pbmc5k_DEV)[["z"]]) %>% c)
head(z_score_mat)
```

### Generate the seed cell index (using the top 5% if too many cells are eligible)
```{r, message = TRUE, warning = FALSE, cache = TRUE}
seed_idx <- seedindex(z_score_mat$z_score, 0.05)
```
**calculate scale factor**
```{r, message = TRUE, warning = FALSE, cache = TRUE}
scale_factor <- cal_scalefactor(z_score=z_score_mat$z_score, 0.01)
```

### Construct m-knn graph
**Calculate tfidf-mat**
```{r, message = TRUE, warning = FALSE, cache = TRUE}
peak_by_cell_mat <- assay(SE_pbmc5k)
tfidf_mat <- tfidf(bmat=peak_by_cell_mat, mat_binary=TRUE, TF=TRUE, log_TF=TRUE)
```
**Calculate lsi-mat**
```{r, message = TRUE, warning = FALSE, cache = TRUE}
lsi_mat <- do_lsi(tfidf_mat, dims=30)
```
**Calculate m-knn graph**
```{r, message = FALSE, warning = FALSE, cache = TRUE}
mutualknn30 <- getmutualknn(lsi_mat, 30)
```

### Network propagation  
```{r, message = FALSE, warning = FALSE, cache = TRUE}
np_score <- randomWalk_sparse(intM=mutualknn30, rownames(mutualknn30)[seed_idx], gamma=0.05)
```
**Trait relevant score (TRS) with scaled and normalized**  
A few cells are singletons are removed from further analysis
```{r, message = FALSE, warning = FALSE, cache = TRUE}
omit_idx <- np_score==0
sum(omit_idx)
mutualknn30 <- mutualknn30[!omit_idx, !omit_idx]
np_score <- np_score[!omit_idx]
TRS <- np_score %>% capOutlierQuantile(., 0.95) %>% max_min_scale
TRS <- TRS * scale_factor
mono_mat <- data.frame(z_score_mat[!omit_idx, ], seed_idx[!omit_idx], np_score, TRS)
head(mono_mat)
```

### UMAP plots of cell type annotation and cell-to-cell graph 
**Cell type annotation**
```{r, message = FALSE, warning = FALSE, cache = TRUE, fig1, fig.height = 3, fig.width = 5, fig.align = "center"}
p <- ggplot(data=mono_mat, aes(x, y, color=color)) + geom_point(size=1, na.rm = TRUE) + 
    pretty_plot() + theme(legend.title = element_blank()) + xlab("UMAP 1") + ylab("UMAP 2")
p
```

**Visualize cell-to-cell graph if you have low-dimensional coordinates such as UMAP1 and UMAP2**
```{r, message = FALSE, warning = FALSE, cache = TRUE, fig0, fig.height = 3, fig.width = 5, fig.align = "center"}
mutualknn30_graph <- graph_from_adjacency_matrix(mutualknn30, mode = "undirected", diag = F)
plot.igraph(mutualknn30_graph, vertex.size=0.8, vertex.label=NA, vertex.color=adjustcolor("#c7ce3d", alpha.f = 1), vertex.frame.color=NA, 
            edge.color=adjustcolor("#443dce", alpha.f = 1), edge.width=0.3, edge.curved=.5, 
            layout=as.matrix(data.frame(mono_mat$x, mono_mat$y)))
```

### Comparsion before and after SCAVENGE analysis  
- Z score based visualization   
**Scatter plot**
```{r, message = FALSE, warning = FALSE, cache = TRUE, fig2, fig.height = 3, fig.width = 5, fig.align = "center"}
viridis = c("#440154FF", "#472D7BFF", "#3B528BFF", "#2C728EFF", "#21908CFF", "#27AD81FF", "#5DC863FF", "#AADC32FF", "#FDE725FF")
p1 <- ggplot(data=mono_mat, aes(x, y, color=z_score)) + geom_point(size=1, na.rm = TRUE, alpha = 0.6) + 
scale_color_gradientn(colors = viridis) + scale_alpha()+
    pretty_plot() + theme(legend.title = element_blank()) + xlab("UMAP 1") + ylab("UMAP 2")
p1
```
**Bar plot**  
```{r, message = FALSE, warning = FALSE, cache = TRUE, fig3, fig.height = 3, fig.width = 5, fig.align = "center"}
pp1 <- ggplot(data=mono_mat,  aes(x=color, y=z_score))  +
    geom_boxplot(aes(fill=color, color=color), outlier.shape=NA) + 
    guides(fill=FALSE) + pretty_plot(fontsize = 10) +
    stat_summary(geom = "crossbar", width=0.65, fatten=0, color="black", fun.data = function(x){ return(c(y=median(x), ymin=median(x), ymax=median(x))) }) + theme(legend.position = "none")
pp1
```

- SCAVENGE TRS based visualization  
**Scatter plot**  
```{r, message = FALSE, warning = FALSE, cache = TRUE, fig4, fig.height = 3, fig.width = 5, fig.align = "center"}
p2 <- ggplot(data=mono_mat, aes(x, y, color=TRS)) + geom_point(size=1, na.rm = TRUE, alpha = 0.6) + 
scale_color_gradientn(colors = viridis) + scale_alpha()+
    pretty_plot() + theme(legend.title = element_blank()) + xlab("UMAP 1") + ylab("UMAP 2")
p2
```
**Bar plot**
```{r, message = FALSE, warning = FALSE, cache = TRUE, fig5, fig.height = 3, fig.width = 5, fig.align = "center"}
pp2 <- ggplot(data=mono_mat,  aes(x=color, y=TRS))  +
    geom_boxplot(aes(fill=color, color=color), outlier.shape=NA) + 
    guides(fill=FALSE) + pretty_plot(fontsize = 10) +
    stat_summary(geom = "crossbar", width=0.65, fatten=0, color="black", fun.data = function(x){ return(c(y=median(x), ymin=median(x), ymax=median(x))) }) + theme(legend.position = "none")
pp2
```

### Trait relevant cell determination from permutation test 
**About 2 mins**  
please set @mycores >= 1 and @permutation_times >= 1,000 in the real setting
```{r, message = FALSE, warning = FALSE, cache = TRUE}
mono_permu <- get_sigcell_simple(knn_sparse_mat=mutualknn30, seed_idx=mono_mat$seed_idx, topseed_npscore=mono_mat$np_score, permutation_times=1000, true_cell_significance=0.05, rda_output=F, mycores=8, rw_gamma=0.05)
mono_mat2 <- data.frame(mono_mat, mono_permu)
```

### Look at the distribution of statistically significant phenotypically enriched and depleted cells
**Enriched cells**
```{r, message = FALSE, warning = FALSE, cache = TRUE, fig6, fig.height = 3, fig.width = 6, fig.align = "center"}
mono_mat2 %>%
    group_by(color) %>% 
        summarise(enriched_cell=sum(true_cell_top_idx)) %>% 
            ggplot(aes(x=color, y=enriched_cell, fill=color)) + geom_bar(stat="identity") + theme_classic()
```

**Depleted cells**
```{r, message = FALSE, warning = FALSE, cache = TRUE, fig7, fig.height = 3, fig.width = 5, fig.align = "center"}
mono_mat2$rev_true_cell_top_idx <- !mono_mat2$true_cell_top_idx
mono_mat2 %>%
    group_by(color) %>% 
        summarise(depleted_cell=sum(rev_true_cell_top_idx)) %>% 
            ggplot(aes(x=color, y=depleted_cell, fill=color)) + geom_bar(stat="identity") + theme_classic()
```
